---
title: "ALDA Examples"
output:
  html_document:
    code_folding: show
  pdf_document: default
---

<https://stats.idre.ucla.edu/r/examples/alda/>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, hidden.lines=T)

library(lattice)
library(latticeExtra)
library(plotrix)

```

```{r function-defs, include=FALSE}
# inspired by: https://stackoverflow.com/questions/35627075/r-lattice-multiple-plot-page-how-to-put-text-in-page-margin
suptitle <- function(title) {
  require(grid, quietly=T)
  vp<-viewport(x=0.5,y=1,width=1,height=0.05, just=c('center','top'))
  pushViewport(vp)
  grid.text(title,gp=gpar(cex=1.5))
  popViewport()
}
```

## Chapter 1

Introduces importance of 

* longitudinal data
* 3 or more timepoints
* important qualities of tests
    + equitable across time
    + equal validity (tests same thing. Nice example: multiplication tests math skill in kids, memory in adults)
    + minimizing errors (though not need to be identical: heteroscadistic ok)
  
***

## Chapter 2

#### Data Formats
* person-level  
    (each timepoint is a different variable. Can't manage heterogeneity in spacing, number of TPs, nor time-varying predictors)
    
* person-period  
    (each timepoint on a separate row)
    

#### Empirical Growth Plots

```{r ch2-load-plot, echo=F}

tolerance.pp <- read.table("https://stats.idre.ucla.edu/wp-content/uploads/2016/02/tolerance1_pp.txt", sep=",", header=T)
tolerance.pp$id <- as.factor(tolerance.pp$id) #so id will be displayed in each plot's title
attach(tolerance.pp)

```

```{r ch2-growth-plots}
# points
xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,ylim=c(0,5),type='b')
suptitle('Tolerance Raw Data')

# points + loess -- fails with singularities                  #@hide@
#xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,  #@hide@
#       type=c("p","smooth")) #point and loess                #@hide@

#points + regression
xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,ylim=c(0,5),
       type=c("p","r")) #point and regression line
suptitle('Tolerance OLS Fits')

#xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,    #@hide@
#      par.settings = theEconomist.theme(box = "transparent"),  #@hide@
#      lattice.options = theEconomist.opts())                   #@hide@
#@hide@
#asTheEconomist(xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T))  #@hide@
```

#### Fitting OLS regression model for each person

```{r ch2-subject-fits, attr.output = 'style="max-height: 500px;"'}
by(tolerance.pp, id, function(x) summary(lm(tolerance ~ time, data=x)))
```

#### Table 2.2, Figure 2.4
```{r ch2-individual-coeffs}
# recreate Table 2.2

#int <- by(tolerance.pp, id, function(x) #UCLA way
#  coefficients(lm(tolerance ~ time, data=x))[[1]])

#more compact way
fits <- by(tolerance.pp, id, function(x) lm(tolerance ~ time, data=x) )
coefs = sapply(fits, coef)
se = sapply(fits, function(x) summary(x)$coefficients[,'Std. Error'])

print("Initial Status")
summary(coefs['(Intercept)',])
stem(coefs['(Intercept)',], scale=2)

print("Rate of Change")
summary(coefs['time',])
stem(coefs['time',], scale=2)

print('Residial variance')
rv = sapply(fits, function(x) summary(x)$sigma**2 )
summary(rv)

print("R^2")
rsq = sapply(fits,function(x) summary(x)$r.squared )
summary(rsq)



```

```{r interaction-plot}
fit.vals = sapply(fits,fitted.values)
allfit <- lm(tolerance~time,data=tolerance.pp)

with(tolerance.pp,
     interaction.plot(age,id,tolerance,ylab='tolerance'))
ablineclip(reg=allfit, col='purple', lwd=4,x1=1,x2=5)
legend(1,3.5,c('individuals','mean'),col=c('black','purple'),lwd=c(1,4))

with(tolerance.pp,
     interaction.plot(age,id,fit.vals,ylab="fitted tolerance"))
ablineclip(a=coef(allfit)[1], b=coef(allfit)[2], col='purple', lwd=4, x1=1,x2=5)
legend(1,3.5,c('individuals','mean'),col=c('black','purple'),lwd=c(1,4))

```

### Aside Question
The UCLA source finds a mean fit by fitting all of the subjects data. How does this differ from taking means of coefficients?

A: Not at all. It would surely be different if there was e.g. missing data

```{r}
allfit <- lm(tolerance~time,data=tolerance.pp)
coef(allfit)

rowMeans(coefs)
```

***

## Chapter 5 on TIME


```{r}
reading <- read.table("https://stats.idre.ucla.edu/stat/r/examples/alda/data/reading_pp.txt", header=T, sep=",")
reading[reading$id %in% c(4, 27, 31, 33, 41, 49, 69, 77, 87), ]
summary(reading)
```

## Empirical Change Plots


```{r}
library(lattice)

xyplot(piat~age | id,
       data=reading[reading$id %in% c(4, 27, 31, 33, 41, 49, 69, 77, 87), ],
       panel=function(x,y,subscripts) {
         panel.xyplot(x,y,pch=16)
         panel.lmline(x,y,lty=1)
         panel.xyplot(reading$agegrp[subscripts],y,pch=3)
         panel.lmline(reading$agegrp[subscripts],y,lty=4) 
         },
       xlab='age or agegrp',ylim=c(0,80), as.table=T, subscripts=T, 
       key=list(columns=1,
                text=list(lab=c('age','agegrp')),
                points=list(pch=c(16,3)),
                lines=list(lty=c(1,4))
                ) 
       )
       
```

