---
title: "ALDA Examples"
output:
  html_document:
    code_folding: show
  pdf_document: default
---

<https://stats.idre.ucla.edu/r/examples/alda/>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE)

library(lattice)
library(latticeExtra)
library(grid)
library(plotrix)
library(tidyverse)
library(magrittr)
library(nlme)
library(lme4)

```

```{r function-defs, include=FALSE}
# inspired by: https://stackoverflow.com/questions/35627075/r-lattice-multiple-plot-page-how-to-put-text-in-page-margin
suptitle <- function(title, cex=1.5) {
  require(grid, quietly=T)
  vp<-viewport(x=0.5,y=1,width=1,height=0.05, just=c('center','top'))
  pushViewport(vp)
  grid.text(title,gp=gpar(cex=cex))
  popViewport()
}
```


# Chapter 4

Now things start getting really good! Combine level-1 and level-2 models into a single model that
has several forms of expression that are really informative

### Composite formulation

This shows clearly what's meant by random effect of subject for intercept ($\zeta_{0i}$) 
and slope ($\zeta_{1i}$)...that which measures the entirety of 
the deviation of each subject $i$ from the population mean intercept and slope $\gamma_{00}$ and $\gamma_{10}$

#### CONDITION = 0
$$Y_{ij} = (\gamma_{00} + \zeta_{0i}) + (\gamma_{10} + \zeta_{1_i}) * TIME_{ij} + \epsilon_{ij}$$

#### CONDITION = 1

$$Y_{ij} = (\gamma_{00} + \gamma_{01} + \zeta_{0i}) + (\gamma_{10} + \gamma_{11} + \zeta_{1_i}) * TIME_{ij} + \epsilon_{ij}$$

#### Together

$$Y_{ij} = 
\underbrace{
(\gamma_{00} + \gamma_{01}*CONDITION_i + \zeta_{0i})
}_{\pi_{0i}}
+ 
\underbrace{
(\gamma_{10} + \gamma_{11}*CONDITION_i + \zeta_{1_i})
}_{\pi_{1i}}
* TIME_{ij} 
+ \epsilon_{ij}
$$


### Fixed + rando0im formulation

Rearranged, this clearly distinguishes fixed and radom effects portions of the model

$$Y_{ij} = 
\underbrace{
\left[ \gamma_{00} + \gamma_{01} * CONDITION_i + \gamma_{10} * TIME_{ij} + 
    \gamma_{11} * (CONDITION_i \times TIME_{ij}) \right]
}_{fixed} 
+ 
\underbrace{
\left[
\zeta_{0i} + \zeta_{1i} * TIME_{ij} + \epsilon_{ij}
\right]
}_{random}
$$

### random effect distributions

As before:

$$\epsilon_{ij} \sim N(0, \sigma_\epsilon^2) \\ 
\\
\begin{bmatrix}
\zeta_{0i} \\ \zeta_{1i}
\end{bmatrix}
\sim \mathcal{N}
\left(
\begin{bmatrix}
  0 \\ 0
\end{bmatrix},
\begin{bmatrix}
  \sigma_{0}^2 & \sigma_{01} \\ \sigma_{10} & \sigma_{1}^2
\end{bmatrix}
\right)$$

## Examples

```{r load data}
alcohol1 <- read.table("https://stats.idre.ucla.edu/stat/r/examples/alda/data/alcohol1_pp.txt", header=T, sep=",")
alcohol1$id %<>% as.factor
if (! 'alcohol1' %in% search()) {
  attach(alcohol1)
}
```

## Empirical Growth Plots

```{r plot}
subset <- alcohol1[alcohol1$id %in% c(4, 14, 23, 32, 41, 56, 65, 82), ]
subset

xyplot( alcuse ~ age | id,
        data = subset,
        as.table=T, ylim=c(-1,4),type=c("p","r"))
suptitle('Fig 4.1, selected empirical growth plots + OLS fits')
```


```{r ucla interaction plot by coa}
alcohol.coa0 <- alcohol1 %>% filter(coa==0) %>% droplevels()
fits.coa0 <- by(alcohol.coa0, alcohol.coa0$id,
               function(data) lm(alcuse~age, data=data))

interaction.plot(alcohol.coa0$age, alcohol.coa0$id, sapply(fits.coa0,fitted.values),
                 xlab='AGE',ylab='ALCUSE', ylim=c(-1,4), lwd=1)
suptitle('Fig 4.2 OLS Fit, Alcuse vs Age, coa=0')
  
```


```{r fit individual models, echo=FALSE}
fits <- by(alcohol1, id, function(x) lm(alcuse ~ age, data=x) )
coefs = sapply(fits, coef)
fit.vals = sapply(fits,fitted.values)
se = sapply(fits, function(x) summary(x)$coefficients[,'Std. Error'])
rv = sapply(fits, function(x) summary(x)$sigma**2 )
rsq = sapply(fits,function(x) summary(x)$r.squared )
```

```{r interaction plots}
ggplot(alcohol1, aes(age, fit.vals, group=id, color=factor(coa))) +
  geom_line() +
  scale_colour_manual(values = c('blue','red')) +
  stat_smooth(aes(group=factor(coa), fill=factor(coa)), method='lm') +
  facet_wrap(~coa)
```

## Models

The book takes the approach of building successively more complete models. It suggests first fitting 
Unconditional Means and Unconditional Growth models, literally with no $CONDITION$ dependent variable.

### Unconditional Means Model

$$Y_{ij} = \pi_{0i} + \epsilon_{ij} \\ \pi_{0i} = \gamma_{00} + \zeta_{0i} \\
Y_{ij} = \left[  \gamma_{00}  \right] + \left[ \epsilon_{ij} + \zeta_{0i} \right]
$$

```{r fit model A}
model.a <- lme(alcuse ~ 1, alcohol1, random = ~1 | id)
summary(model.a)
```

#### Evaluation


#### Compare with lme4::lmer
```{r fit model A lmer}
model.a2 <- lmer(alcuse ~ 1 + (1 | id), alcohol1)
summary(model.a2)
```
Interesting note on why lmer does not include p-values:
https://stat.ethz.ch/pipermail/r-help/2006-May/094765.html


### Unconditional Growth Model

$$Y_{ij} = \pi_{0i} + \pi_{1i}*TIME_{ij} + \epsilon_{ij} \\ 
\pi_{0i} = \gamma_{00} + \zeta_{0i} \\
\pi_{1i} = \gamma_{10} + \zeta_{1i} \\
Y_{ij} = \left[  \gamma_{00} + \gamma_{10} * TIME_{ij} \right] + 
\left[ \epsilon_{ij} + \zeta_{0i} + \zeta_{1i} * TIME_{ij}  \right]
$$

```{r fit model B}
model.b <- lme(alcuse ~ 1 + age_14, alcohol1, random = ~ 1 + age_14 | id, method = 'ML') #1 are implied if omitted
summary(model.b)
```

## Interpretation


### Plotting Fits
```{r fits, eval=F}
a <- fitted.values(model1)
interaction.plot(age, program, a, xlab="AGE", ylab="COG", 
                 ylim=c(50, 150), lwd=4, lty=1, col=4:5)

```

