---
title: "R Notebook"
output: html_notebook
---

<https://stats.idre.ucla.edu/r/examples/alda/>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = FALSE, hidden.lines=T)

library(lattice)
library(latticeExtra)
library(plotrix)

```

```{r function-defs, include=FALSE}
# inspired by: https://stackoverflow.com/questions/35627075/r-lattice-multiple-plot-page-how-to-put-text-in-page-margin
suptitle <- function(title) {
  require(grid, quietly=T)
  vp<-viewport(x=0.5,y=1,width=1,height=0.05, just=c('center','top'))
  pushViewport(vp)
  grid.text(title,gp=gpar(cex=1.5))
  popViewport()
}
```

## Chapter 1

Introduces importance of 

* longitudinal data
* 3 or more timepoints
* important qualities of tests
    + equitable across time
    + equal validity (tests same thing. Nice example: multiplication tests math skill in kids, memory in adults)
    + minimizing errors (though not need to be identical: heteroscadistic ok)
  
***

## Chapter 2

#### Data Formats
* person-level  
    (each timepoint is a different variable. Can't manage heterogeneity in spacing, number of TPs, nor time-varying predictors)
    
* person-period  
    (each timepoint on a separate row)
    

#### Empirical Growth Plots

```{r ch2-load-plot, echo=F}

tolerance.pp <- read.table("https://stats.idre.ucla.edu/wp-content/uploads/2016/02/tolerance1_pp.txt", sep=",", header=T)
tolerance.pp$id <- as.factor(tolerance.pp$id) #so id will be displayed in each plot's title
attach(tolerance.pp)

```

```{r ch2-growth-plots}
# points
xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,ylim=c(0,5),type='b')
suptitle('Tolerance Raw Data')

# points + loess -- fails with singularities                  #@hide@
#xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,  #@hide@
#       type=c("p","smooth")) #point and loess                #@hide@

#points + regression
xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,ylim=c(0,5),
       type=c("p","r")) #point and regression line
suptitle('Tolerance OLS Fits')

#xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T,    #@hide@
#      par.settings = theEconomist.theme(box = "transparent"),  #@hide@
#      lattice.options = theEconomist.opts())                   #@hide@
#@hide@
#asTheEconomist(xyplot(tolerance ~ age | id, data=tolerance.pp, as.table=T))  #@hide@
```
